Preferences for market justice: the role of intergenerational social mobility and beliefs in meritocracy in Chile

### 2025-05-21 advances/informe2.tex
Update

Since the early 1980s, many countries have experienced a widespread
retreatment of universal welfare programs and have adopted a trend
toward privatization and commoditization of various public goods,
welfare policies and social services
(\citeproc{ref-gingrich_making_2011}{Gingrich, 2011};
\citeproc{ref-salamon_marketization_1993}{Salamon, 1993};
\citeproc{ref-streeck_how_2016}{Streeck, 2016}). Neoliberal reforms
reshaped the architecture of welfare-state institutions across Western
democracies and Latin America, extending market criteria into areas of
social reproduction that were traditionally governed by the state. This
shift diminished the role of public provision of core social services
(\citeproc{ref-arrizabalo_milagro_1995}{Arrizabalo, 1995};
\citeproc{ref-busemeyer_welfare_2020}{Busemeyer \& Iversen, 2020}).
Echoing Polanyi's (\citeproc{ref-polanyi_great_1975}{1975}) insight that
markets constitute a distinct moral order, the institutional diffusion
of market rules has fostered a corresponding moral economy---a
constellation of norms and values about fair allocation embedded in, and
reinforced by, social institutions (\citeproc{ref-mau_moral_2014}{Mau,
2014}; \citeproc{ref-svallforsMoralEconomyClass2006a}{Svallfors, 2006}).
Within this moral economy, a growing body of research has been addressed
the extent to which, and the mechanisms through which, citizens deem it
just that access to social services such as health care, pensions, and
education be determined by market-based criteria, a phenomenon labelled
market justice preferences
(\citeproc{ref-busemeyer_skills_2014}{Busemeyer, 2014};
\citeproc{ref-castillo_socialization_2024}{Castillo et al., 2024};
\citeproc{ref-immergut_it_2020}{Immergut \& Schneider, 2020};
\citeproc{ref-koos_moral_2019}{Koos \& Sachweh, 2019};
\citeproc{ref-lindh_public_2015}{Lindh, 2015};
\citeproc{ref-lindh_bringing_2023}{Lindh \& McCall, 2023}).
Understanding these preferences is crucial because they can legitimize
economic inequality by framing it as the natural outcome of fair
competition, driven by individual responsibility and limited state
intervention (\citeproc{ref-svallfors_political_2007}{Svallfors, 2007}).

Existing literature consistently demonstrates that market justice
preferences are shaped by both the economic and institutional contexts
of countries and individuals' positions within social stratification.
Recent studies indicate that in countries with greater public provision
of social services or more extensive welfare states, market justice
preferences tend to be lower
(\citeproc{ref-busemeyer_skills_2014}{Busemeyer, 2014};
\citeproc{ref-immergut_it_2020}{Immergut \& Schneider, 2020}), whereas
in contexts with higher levels of privatization, preferences for
market-based criteria are more pronounced
(\citeproc{ref-castillo_socialization_2024}{Castillo et al., 2024};
\citeproc{ref-lindh_public_2015}{Lindh, 2015}). In such contexts, market
justice preferences tend to rise as individuals ``ascend'' the social
structure, with those in more privileged positions in terms of class,
education, and income holding stronger preferences for market-based
solutions compared to those in more disadvantaged or at-risk positions
(\citeproc{ref-busemeyer_skills_2014}{Busemeyer, 2014};
\citeproc{ref-immergut_it_2020}{Immergut \& Schneider, 2020};
\citeproc{ref-kluegel_legitimation_1999}{Kluegel et al., 1999};
\citeproc{ref-lindh_public_2015}{Lindh, 2015};
\citeproc{ref-svallfors_political_2007}{Svallfors, 2007}). Empirically,
it has been consistently demonstrated that those in socioeconomically
advantaged positions endorse the idea that those with higher incomes
should be able to pay more for better social services in the domains of
education (\citeproc{ref-lee_fairness_2023}{Lee \& Stacey, 2023}),
healthcare (\citeproc{ref-immergut_it_2020}{Immergut \& Schneider,
2020}; \citeproc{ref-vondemknesebeck_are_2016}{Von Dem Knesebeck et al.,
2016}) and old age pensions
(\citeproc{ref-castillo_socialization_2024}{Castillo et al., 2024};
\citeproc{ref-otero_power_2024}{Otero \& Mendoza, 2024}).

Although it is clear that one's social position influences market
preferences, the question of how upward or downward mobility within the
social structure affects these preferences remains unanswered. This is
far from trivial, as many individuals in Western democracies and Latin
America have experienced varying degrees of social mobility
(\citeproc{ref-breen_social_2004}{Breen, 2004};
\citeproc{ref-lopez-roldan_comparative_2021}{López-Roldán \& Fachelli,
2021}). The social origins and destinations of individuals exert
different effects on a range of attitudes
(\citeproc{ref-gugushvili_trends_2014}{Gugushvili, 2014}), while
movement between these positions exposes individuals to distinct
processes and mechanisms that can profoundly influence their
perspectives, shaping how they perceive the world and evaluate what they
consider fair (\citeproc{ref-gugushvili_trends_2014}{Gugushvili,
2014})). Some of the current research on attitudes toward inequality has
concluded that the type of social mobility---upward or downward---that
individuals experience, whether measured objectively or subjectively,
has differentiated effects on support for redistribution
(\citeproc{ref-alesina_intergenerational_2018}{Alesina et al., 2018};
\citeproc{ref-benabou_social_2001}{Benabou \& Ok, 2001};
\citeproc{ref-jaime-castillo_social_2019}{Jaime-Castillo \&
Marqués-Perales, 2019}; \citeproc{ref-schmidt_experience_2011}{Schmidt,
2011}), system legitimacy, and meritocratic beliefs
(\citeproc{ref-day_movin_2017}{Day \& Fiske, 2017};
\citeproc{ref-mijs_belief_2022}{Mijs et al., 2022}), as well as on
tolerance for economic inequality
(\citeproc{ref-gugushvili_intergenerational_2016c}{Gugushvili, 2016a};
\citeproc{ref-shariff_income_2016}{Shariff et al., 2016}), attributions
about poverty and wealth (\citeproc{ref-bucca_merit_2016}{Bucca, 2016};
\citeproc{ref-gugushvili_intergenerational_2016}{Gugushvili, 2016b}),
and support for welfare policies and market economies
(\citeproc{ref-gugushvili_subjective_2017}{Gugushvili, 2017};
\citeproc{ref-wen_does_2021}{Wen \& Witteveen, 2021}). Thus, building on
these contributions, expanding the analysis of the effects of social
mobility on market justice preferences could contribute to this emerging
literature and shed light on the mechanisms behind the justification of
inequalities in access to social services.

Market justice preferences are shaped not only by objective
socioeconomic factors but also by perceptions and beliefs about
meritocracy. Meritocracy asserts that inequality is an inherent feature
of societies but can be legitimized through principles such as effort
and talent (\citeproc{ref-davis_principles_2001}{Davis \& Moore, 2001};
\citeproc{ref-young_rise_1958}{Young, 1958}). Previous studies have
shown that individuals with stronger meritocratic beliefs tend to
perceive less inequality, attributing economic differences to individual
achievements (\citeproc{ref-mijs_paradox_2019}{Mijs, 2019};
\citeproc{ref-wilson_role_2003}{Wilson, 2003}). Additionally, these
beliefs justify greater inequality, as they are associated with
attitudes that uphold the legitimization of social differences
(\citeproc{ref-batruch_belief_2023}{Batruch et al., 2023}). In unequal
societies where the distribution of goods and services is predominantly
governed by market logic, such beliefs can play a critical role in the
acceptance of social inequalities. Recent research by Castillo et al.
(\citeproc{ref-castillo_socialization_2024}{2024}) demonstrates that
students in Chile who believe that effort and talent are rewarded in
their country hold stronger preferences for market justice in access to
healthcare, pensions, and education. By incorporating perceptions of
meritocracy into this study, we can better understand how preferences
toward market justice are shaped not only by an individual's structural
social position but also by beliefs about the justification of these
inequalities. Moreover, these beliefs are often solidified early in
life, reinforced by institutions that promote values such as effort and
individual skills as means to upward social mobility
(\citeproc{ref-castillo_socialization_2024}{Castillo et al., 2024};
\citeproc{ref-reynolds_perceptions_2014}{Reynolds \& Xian, 2014}).

Beyond the isolated effects of social mobility and beliefs in
meritocracy on market justice preferences, these factors interact to
shape these preferences in a complex way. One proposed mechanism in the
literature regarding the process of social mobility that leads to
differences in distributive justice preferences is the psychological
mechanism of self-serving bias in causal attribution
(\citeproc{ref-gugushvili_intergenerational_2016c}{Gugushvili, 2016a};
\citeproc{ref-schmidt_experience_2011}{Schmidt, 2011}). Self-serving
bias suggests that people attribute failures, such as downward mobility,
to factors beyond their control, while explaining successes, such as
upward mobility, by pointing to their own merits and effort
(\citeproc{ref-miller_selfserving_1975}{Miller \& Ross, 1975}).
Individuals who experience upward mobility tend to view their position
as a result of their own effort and skills, making them more likely to
support the idea that individuals are responsible for their own success
or failure. Research shows that upward mobility is associated with lower
redistribution preferences
(\citeproc{ref-gugushvili_intergenerational_2016c}{Gugushvili, 2016a};
\citeproc{ref-schmidt_experience_2011}{Schmidt, 2011}) and stronger
legitimacy of income inequality
(\citeproc{ref-shariff_income_2016}{Shariff et al., 2016}). In contrast,
individuals who experience downward mobility tend to attribute their
failure to external factors, such as inequality, and are more supportive
of redistribution and the idea that individuals should not be blamed for
their economic failure
(\citeproc{ref-gugushvili_intergenerational_2016c}{Gugushvili, 2016a};
\citeproc{ref-schmidt_experience_2011}{Schmidt, 2011}). Taken together,
I suggest that meritocratic beliefs may mirror this self-serving bias
mechanism, as they reinforce the tendency to justify one's social status
as the result of personal merits, which is closely related to the
attribution bias.

Focusing in Chile, this article aims to achieve two main objectives:
first, to analyze the extent to which intergenerational social mobility
affects market justice preferences on healthcare, pensions and
education, and second, to examine how meritocratic beliefs may moderate
this relationship. The central argument is that upward social mobility
will have a positive effect on market justice preferences, while
downward mobility will have a negative effect. Additionally,
meritocratic beliefs will moderate this relationship by reflecting a
self-serving bias mechanism of causal attribution.

The Chilean case is particularly intriguing for studying market justice
preferences. Despite being one of the most prosperous nations in Latin
America, it has one of the highest levels of economic inequality in the
region. The poorest 50\% of Chileans capture only 10\% of the total
income and hold negative wealth, while the richest 1\% receives nearly
27\% of the income and controls 49.6\% of the wealth
(\citeproc{ref-chancel_world_2022}{Chancel et al., 2022}). This economic
inequality exists alongside notable social mobility, with significant
upward mobility in recent decades
(\citeproc{ref-espinoza_estratificacion_2013}{Espinoza et al., 2013};
\citeproc{ref-torche_intergenerational_2014}{Torche, 2014}). Chile's
social policy regime is characterized by a welfare model that relies
heavily on private provision in healthcare, pensions, and education
systems, often segmented by individuals' ability to pay and heavily
dependent on state subsidies (\citeproc{ref-boccardo_30_2020}{Boccardo,
2020}).

Against this backdrop, the main research question is: To what extent
does intergenerational social mobility affect market justice
preferences, and how do meritocratic beliefs explain this relationship
in Chile? This study aims to contribute to the literature by providing
evidence from a Latin American developing country, highlighting how
changes in individuals' socioeconomic position and meritocratic beliefs
influence market justice preferences in critical areas like healthcare,
pensions, and education.

@@

@@

\bibitem[\citeproctext]{ref-benabou_social_2001}
Benabou, R., \& Ok, E. A. (2001). Social {Mobility} and the {Demand} for
{Redistribution}: {The Poum Hypothesis}. \emph{The Quarterly Journal of
Economics}, \emph{116}(2), 447--487.
\url{https://doi.org/10.1162/00335530151144078}

@@

\bibitem[\citeproctext]{ref-breen_social_2004}
Breen, R. (2004). \emph{Social {Mobility} in {Europe}}. OUP Oxford.

@@

\bibitem[\citeproctext]{ref-bucca_merit_2016}
Bucca, M. (2016). Merit and blame in unequal societies: {Explaining
Latin Americans}' beliefs about wealth and poverty. \emph{Research in
Social Stratification and Mobility}, \emph{44}, 98--112.
\url{https://doi.org/10.1016/j.rssm.2016.02.005}

@@

\bibitem[\citeproctext]{ref-busemeyer_welfare_2020}
Busemeyer, M. R., \& Iversen, T. (2020). The {Welfare State} with
{Private Alternatives}: {The Transformation} of {Popular Support} for
{Social Insurance}. \emph{The Journal of Politics}, \emph{82}(2),
671--686. \url{https://doi.org/10.1086/706980}

@@

@@

@@

@@

\bibitem[\citeproctext]{ref-kluegel_legitimation_1999}
Kluegel, J. R., Mason, D. S., \& Wegener, B. (1999). The {Legitimation}
of {Capitalism} in the {Postcommunist Transition}: {Public Opinion}
about {Market Justice}, 1991-1996. \emph{European Sociological Review},
\emph{15}(3), 251--283. Retrieved from
\url{https://www.jstor.org/stable/522731}

@@

\bibitem[\citeproctext]{ref-mau_moral_2014}
Mau, S. (2014). \emph{The moral economy of welfare states: {Britain} and
{Germany} compared} (First issued in paperback). London New York NY:
Routledge \& Francis Group.

@@

\bibitem[\citeproctext]{ref-reynolds_perceptions_2014}
Reynolds, J., \& Xian, H. (2014). Perceptions of meritocracy in the land
of opportunity. \emph{Research in Social Stratification and Mobility},
\emph{36}, 121--137. \url{https://doi.org/10.1016/j.rssm.2014.03.001}

\bibitem[\citeproctext]{ref-salamon_marketization_1993}
Salamon, L. M. (1993). The {Marketization} of {Welfare}: {Changing
Nonprofit} and {For-Profit Roles} in the {American Welfare State}.
\emph{Social Service Review}, \emph{67}(1), 16--39.
\url{https://doi.org/10.1086/603963}

@@

>>>>

What makes market-driven inequalities appear legitimate in the eyes of
the public? Since the early 1980s, many countries have experienced a
widespread retreat from universal welfare programs and a shift toward
the privatization and commodification of public goods, welfare policies,
and social services (\citeproc{ref-gingrich_making_2011}{Gingrich,
2011}; \citeproc{ref-streeck_how_2016}{Streeck, 2016}). In Latin
America, as elsewhere, neoliberal reforms reshaped welfare-state
institutions by extending market logic into domains of social
reproduction that were traditionally governed by the state
(\citeproc{ref-arrizabalo_milagro_1995}{Arrizabalo, 1995};
\citeproc{ref-ferre_welfare_2023}{Ferre, 2023}). This transformation
reduced the role of public provision and increased the presence of
private actors in core social services
(\citeproc{ref-harvey_breve_2015}{Harvey, 2015}). Echoing Polanyi's
(\citeproc{ref-polanyi_great_1975}{1975}) insight that markets
constitute a distinct moral order, the institutional diffusion of market
rules has fostered a corresponding moral economy: a constellation of
norms and values concerning fair allocation, embedded in institutions
and shaping individual subjectivities
(\citeproc{ref-mau_inequality_2015}{Mau, 2015};
\citeproc{ref-svallforsMoralEconomyClass2006a}{Svallfors, 2006}). Within
this framework, a growing body of research examines the extent to which,
and the mechanisms by which, citizens consider it fair that access to
services like health care, pensions, and education be governed by
market-based criteria---a phenomenon known as \emph{market justice
preferences} (\citeproc{ref-busemeyer_skills_2014}{Busemeyer, 2014};
\citeproc{ref-castillo_socialization_2024}{Castillo et al., 2024};
\citeproc{ref-immergut_it_2020}{Immergut \& Schneider, 2020};
\citeproc{ref-koos_moral_2019}{Koos \& Sachweh, 2019};
\citeproc{ref-lindh_public_2015}{Lindh, 2015};
\citeproc{ref-lindh_bringing_2023}{Lindh \& McCall, 2023}).
Understanding these preferences is crucial, as they contribute to
legitimizing economic inequality by framing it as the fair result of
individual responsibility and limited state intervention
(\citeproc{ref-mau_inequality_2015}{Mau, 2015}).

Existing literature shows that market justice preferences are shaped by
both the economic and institutional context and individuals' positions
within social stratification. Grounded in the notion that institutions
influence individual attitudes
(\citeproc{ref-jacksonActorsInstitutions2010}{Jackson, 2010}), studies
find that countries with stronger public provision or more expansive
welfare states exhibit lower levels of market justice preferences
(\citeproc{ref-busemeyer_skills_2014}{Busemeyer, 2014};
\citeproc{ref-immergut_it_2020}{Immergut \& Schneider, 2020}), while
more privatized contexts show stronger support for market-based criteria
(\citeproc{ref-castillo_socialization_2024}{Castillo et al., 2024};
\citeproc{ref-lindh_public_2015}{Lindh, 2015}). In such contexts, market
justice preferences tend to rise as individuals ``ascend'' the social
structure, with those in more privileged positions in terms of class,
education, and income holding stronger preferences for market-based
solutions compared to those in more disadvantaged or at-risk positions
(\citeproc{ref-castillo_socialization_2024}{Castillo et al., 2024};
\citeproc{ref-immergut_it_2020}{Immergut \& Schneider, 2020};
\citeproc{ref-lee_fairness_2023}{Lee \& Stacey, 2023};
\citeproc{ref-lindh_public_2015}{Lindh, 2015};
\citeproc{ref-otero_power_2024}{Otero \& Mendoza, 2024};
\citeproc{ref-svallfors_political_2007}{Svallfors, 2007};
\citeproc{ref-vondemknesebeck_are_2016}{Von Dem Knesebeck et al.,
2016}).

Market justice preferences are shaped not only by objective
socioeconomic conditions but also by popular beliefs about inequality.
Among these, meritocracy is a key normative principle underpinning
market-based distributive preferences
(\citeproc{ref-mau_inequality_2015}{Mau, 2015}). It frames inequality as
inevitable but justifiable through effort and talent
(\citeproc{ref-davis_principles_2001}{Davis \& Moore, 2001};
\citeproc{ref-young_rise_1958}{Young, 1958}). Studies show that
individuals with stronger meritocratic beliefs tend to perceive less
inequality and legitimize it by attributing economic differences to
personal achievement (\citeproc{ref-batruch_belief_2023}{Batruch et al.,
2023}; \citeproc{ref-mijs_paradox_2019}{Mijs, 2019};
\citeproc{ref-wilson_role_2003}{Wilson, 2003}). In highly unequal
societies where access to services is largely governed by market logic,
such beliefs play a critical role in normalizing inequality. Recent
evidence from Chile shows that students who believe effort and talent
are rewarded in their country express stronger preferences for
market-based access to healthcare, pensions, and education
(\citeproc{ref-castillo_socialization_2024}{Castillo et al., 2024}).

Although it is clear that one's social position influences market
preferences, the question of how upward or downward mobility within the
social structure affects these preferences remains unanswered. This
question is far from trivial, especially in Latin America, where many
have experienced various forms of mobility amid high economic inequality
and deep welfare privatization (\citeproc{ref-ferre_welfare_2023}{Ferre,
2023}; \citeproc{ref-lopez-roldan_comparative_2021}{López-Roldán \&
Fachelli, 2021}; \citeproc{ref-torche_intergenerational_2014}{Torche,
2014}). Social origins and destinations affect attitudes toward
inequality in distinct ways (\citeproc{ref-day_movin_2017}{Day \& Fiske,
2017}; \citeproc{ref-gugushvili_intergenerational_2016}{Gugushvili,
2016b}, \citeproc{ref-gugushvili_subjective_2017}{2017};
\citeproc{ref-jaime-castillo_social_2019}{Jaime-Castillo \&
Marqués-Perales, 2019}; \citeproc{ref-mijs_belief_2022}{Mijs et al.,
2022}; \citeproc{ref-wen_does_2021}{Wen \& Witteveen, 2021}), while
movement between these positions exposes individuals to different
experiences and mechanisms that shape their views on what is fair
(\citeproc{ref-gugushvili_trends_2014}{Gugushvili, 2014};
\citeproc{ref-mau_inequality_2015}{Mau, 2015}). Building on this
research, examining the effects of social mobility on market justice
preferences can help to illuminate how inequalities in access to social
services are justified among individuals who have experienced, or not,
changes in their social standing, and what are the normative mechanisms
that guide this justification (\citeproc{ref-mau_inequality_2015}{Mau,
2015}).

Beyond their isolated effects, social mobility and meritocratic beliefs
interact in complex ways to shape market justice preferences. A key
mechanism proposed in the literature to explain how mobility influences
distributive justice preferences is the psychological process of
self-serving bias in causal attribution
(\citeproc{ref-gugushvili_intergenerational_2016c}{Gugushvili, 2016a};
\citeproc{ref-schmidt_experience_2011}{Schmidt, 2011}). This bias
suggests that individuals attribute failures---such as downward
mobility---to external factors, while crediting successes---such as
upward mobility---to their own merit and effort
(\citeproc{ref-miller_selfserving_1975}{Miller \& Ross, 1975}). Those
who experience upward mobility tend to view their social position as
earned, making them more likely to believe that individuals are
responsible for their own success or failure. Research shows that upward
mobility is associated with weaker preferences for redistribution
(\citeproc{ref-alesina_intergenerational_2018}{Alesina et al., 2018};
\citeproc{ref-gugushvili_intergenerational_2016c}{Gugushvili, 2016a};
\citeproc{ref-jaime-castillo_social_2019}{Jaime-Castillo \&
Marqués-Perales, 2019}; \citeproc{ref-schmidt_experience_2011}{Schmidt,
2011}) and stronger legitimacy of income inequality
(\citeproc{ref-shariff_income_2016}{Shariff et al., 2016}). In contrast,
individuals who experience downward mobility tend to blame structural
factors like inequality and are more supportive of redistribution while
rejecting merit-based explanations
(\citeproc{ref-gugushvili_trends_2014}{Gugushvili, 2014}). Taken
together, I argue that meritocratic beliefs may reflect---or
reinforce---this self-serving attribution mechanism by legitimizing
one's social status as the outcome of personal merit, closely tied to
attribution bias.

Against this background, this article pursues two main objectives:
first, to analyze the extent to which intergenerational social mobility
influences market justice preferences regarding healthcare, pensions,
and education; and second, to examine how meritocratic beliefs may
moderate this relationship. Building on a theoretical framework that
emphasizes how neoliberal transformations---particularly through the
privatization and commodification of key areas of social
reproduction---have profoundly reshaped processes of subject formation
(\citeproc{ref-mau_inequality_2015}{Mau, 2015}), the central argument is
that upward mobility increases support for market justice preferences,
while downward mobility decreases it. Moreover, meritocratic beliefs are
expected to moderate this relationship by reflecting a self-serving
attribution bias, whereby individuals justify their social position in
terms of personal merit.

This study focuses on Chile, a particularly intriguing case for
examining market justice preferences. Despite significant economic
growth and poverty reduction, Chile has some of the highest levels of
inequality in Latin America and among OECD countries
(\citeproc{ref-chancel_world_2022}{Chancel et al., 2022};
\citeproc{ref-flores_top_2020}{Flores et al., 2020}). This inequality
coexists with short-range upward mobility among lower-class segments
moving into middle strata, though strong barriers remain to reaching
higher positions (\citeproc{ref-espinoza_estratificacion_2013}{Espinoza
et al., 2013}; \citeproc{ref-torche_intergenerational_2014}{Torche,
2014}). What makes Chile especially salient is that much of this
inequality is rooted in deep neoliberal reforms that institutionalized
the privatization and commodification of key social sectors
(\citeproc{ref-madariaga_three_2020}{Madariaga, 2020}). Introduced
during the dictatorship (1973--1989) and expanded in democracy, these
reforms enabled the unprecedented emergence of markets in health,
pensions, and education, with provision segmented by individuals'
ability to pay and supported by public subsidies
(\citeproc{ref-boccardo_30_2020}{Boccardo, 2020}). In parallel---and
despite waves of protest against inequality and commodification from
2006 to 2019 (\citeproc{ref-somma_no_2021}{Somma et al.,
2021})---Chilean subjectivities have been increasingly shaped by
neoliberal discourses and market logics, influencing their attitudes
toward inequality and welfare distribution
(\citeproc{ref-arteagaaguirre_politicas_2015}{Arteaga Aguirre \& Iñigo
Valderrama, 2015}; \citeproc{ref-castillo_socialization_2024}{Castillo
et al., 2024}).

In this context, the questions that guide this research are as follows:

\begin{enumerate}
\def\labelenumi{(\arabic{enumi})}
\tightlist
\item
  To what extent does intergenerational social mobility influence market
  justice preferences regarding healthcare, pensions, and education in
  Chile?
\item
  How do meritocratic beliefs condition or moderate this relationship in
  the Chilean context?
\end{enumerate}

To address these questions, this study draws on large-scale,
representative survey data collected in 2018 from the urban Chilean
population aged 18 to 75 (n = 2,983). The next section outlines the
theoretical framework linking market justice preferences, social
mobility, and meritocratic beliefs, and proposes a set of hypotheses.
This is followed by a description of the data, variables, and analytical
strategy. The final sections present the empirical findings, offer an
interpretation of the results, and conclude with a discussion of their
implications.

\section{Theoretical and empirical
background}\label{theoretical-and-empirical-background}

\subsection{Market justice
preferences}\label{market-justice-preferences}

\subsection{Social mobility}\label{social-mobility}

\subsection{Meritocracy}\label{meritocracy}

\subsection{The Chilean context}\label{the-chilean-context}

@@

\bibitem[\citeproctext]{ref-arteagaaguirre_politicas_2015}
Arteaga Aguirre, C., \& Iñigo Valderrama, I. (2015). {Pol{í}ticas
sociales, modelo de desarrollo y subjetividad de grupos vulnerables en
Chile}.

@@

@@

@@

@@

@@

\bibitem[\citeproctext]{ref-ferre_welfare_2023}
Ferre, J. C. (2023). Welfare regimes in twenty-first-century {Latin
America}. \emph{Journal of International and Comparative Social Policy},
\emph{39}(2), 101--127. \url{https://doi.org/10.1017/ics.2023.16}

\bibitem[\citeproctext]{ref-flores_top_2020}
Flores, I., Sanhueza, C., Atria, J., \& Mayer, R. (2020). Top {Incomes}
in {Chile}: {A Historical Perspective} on {Income Inequality},
1964--2017. \emph{Review of Income and Wealth}, \emph{66}(4), 850--874.
\url{https://doi.org/10.1111/roiw.12441}

@@

\bibitem[\citeproctext]{ref-harvey_breve_2015}
Harvey, D. (2015). \emph{{Breve historia del neoliberalismo}}. Madrid
(Espa{ñ}a): Ediciones Akal.

@@

\bibitem[\citeproctext]{ref-jacksonActorsInstitutions2010}
Jackson, G. (2010). Actors and institutions. In G. Morgan, J. L.
Campbell, C. Crouch, O. K. Pedersen, \& R. Whitley (Eds.), \emph{The
{Oxford Handbook} of {Comparative Institutional Analysis}} (1st ed.).
Oxford University Press.
\url{https://doi.org/10.1093/oxfordhb/9780199233762.001.0001}

@@

@@

\bibitem[\citeproctext]{ref-madariaga_three_2020}
Madariaga, A. (2020). The three pillars of neoliberalism: {Chile}'s
economic policy trajectory in comparative perspective.
\emph{Contemporary Politics}, \emph{26}(3), 308--329.
\url{https://doi.org/10.1080/13569775.2020.1735021}

\bibitem[\citeproctext]{ref-mau_inequality_2015}
Mau, S. (2015). \emph{Inequality, {Marketization} and the {Majority
Class}: {Why Did} the {European Middle Classes Accept Neo-Liberalism}?}
Houndmills: Palgrave Macmillan.

@@

@@

\bibitem[\citeproctext]{ref-somma_no_2021}
Somma, N. M., Bargsted, M., Disi Pavlic, R., \& Medel, R. M. (2021). No
water in the oasis: The {Chilean Spring} of 2019--2020. \emph{Social
Movement Studies}, \emph{20}(4), 495--502.
\url{https://doi.org/10.1080/14742837.2020.1727737}
